# ggplot2

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
```


Useful references:

* [ggplot2 extensions](https://www.ggplot2-exts.org )
* [Theory of ggplot](http://ggplot2.tidyverse.org/reference/)
* [Themes](https://rstudio-pubs-static.s3.amazonaws.com/3364_d1a578f521174152b46b19d0c83cbe7e.html)

## General syntax

The general syntax of a ggplot2 call:

```{r, eval=FALSE}
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>, 
     position = <POSITION>) +
  <COORDINATE_FUNCTION> +
  <FACET_FUNCTION>
```

There are 7 parameters to specify but you will rarely have to specify all 7 parameters as ggplot has useful defaults.

### geoms

Different types of graphs are called different 'geoms', as they can be thought of geometrical objects used to represent data. E.g.:

* geom_bar = bar graph
* geom_line = line graph
* geom_boxplot = box plot        
* geom_point = scatter plot
* geom_smooth = smooth line fitted to the data

### Layering

Think of ggplot2 as creating a plot using layers. Here is a simple scatter plot: 
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

We can break this down into 'layers'.

This creates an empty graph:
```{r}
ggplot(data=mpg)
```

geom_point() creates a scatter plot on top of the empty base. 'mapping' is an argument you beed to define. It maps variable to a way to show it on graph or 'defines how variables in your dataset are mapped to visual properties'.

Mapping is always paired with aes() and the x and y arguments of aes(), which specify which variable is x and which one is y. 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```


The specific `mapping = aes()` options vary depending on the type of 





## geom_point



